A forest is more than its trees: haplotypes and ancestral recombination graphs

### 2025-05-16 paper/review-responses.tex

\begin{point}{\revref}
 I am a bit confused by the definition of weighted Robinson-Foulds distance at \revref.% on the bottom of page 6. 
I believe that this is more like a Robinson-Foulds \emph{similarity}, as I believe that as defined it should be maximized when the two ARGs are the same, and smaller for more dissimilar ARGs.
\end{point}

\reply{ For Peter.
}

\begin{point}{}
 Is the dissimilarity measure defined in equation \eqref{eqn:wrmse} used anywhere?
\end{point}

\reply{No, the weighted mean square error defined in \eqref{eqn:wrmse} is not shown for our later experiments shown in \figref{fig:dissimilarity}, however it is always computed when using our implementation is \texttt{tscompare}. \revref
}

@@

\begin{point}{}
 What is ``dissimilarity'' as mentioned in the caption of \figref{fig:conceptual_discrepancy}?
\end{point}

\reply{``dissimilarity'' was our original language to describe the non-matched spans, and we must've missed one such reference to the old name. We have now changed the caption in \figref{fig:conceptual_discrepancy} to reflect that change. \revref
}

@@

\begin{point}{\revref}
 I found the discussion around the metric version to be hard to follow and worth expanding.
For example, I believe that the edges of the ARG are what's getting embedded into
$[0, L) \times N \times N$, and the symmetric differences are taken in this space in the obvious way (i.e., without this mapping its unclear what the relative difference of two ARGs is).
There's also a mention of ``dissimilarity'' in this paragraph. I assume that that's the ``symmetrized version of non-matched span'' that was just defined, but it would be good to make that clear. 
\end{point}

\reply{Good point. Again ``dissimilarity'' is the original name we meant to remove, so it has been correctly changed to \emph{non-matched spans}. We have also added more explanation to our argument to improve its clarity.\revref
}

@@

\begin{point}{}
 The performance on inferred ARGs makes clear that this is a difficult problem, and that this haplotype view might be useful conceptually and for methods development, but that as of now we should probably not take these ancestral haplotypes at face value in inferred ARGs. It would be helpful if the authors could provide some intuition for what's going wrong in this case. Presumably the topologies are not inferred well, and many nodes can be extended, which often results in some true and false positives? Another use case I could see would be using this approach to simplify inferred ARGs to make various downstream computations faster (a la \figref{fig:speed_and_edges}). As such, I would be curious to see how the extension algorithm affects the number of edges in inferred ARGs.
\end{point}

\reply{Thanks for this feedback. Yes, if an inferred ARG is incorrect for certain nodes, our method of extend haplotypes exacerbates this issue as extending an incorrectly inferred node extends this incorrect node span even further. Based on our tests in \figref{fig:speed_and_edges} and \tabref{tab:edge-counts}, these results would additionally hold for the inferred ARGs. For this reason, we didn't include additional figures for edge counts and computation time on inferred ARGs.
}

@@

\begin{point}{}
 This is obviously a matter of taste, but I would have found the paper a bit easier to follow if some of the points discussed in the discussion were brought up earlier in the paper. For example, the ``ignorance and omission in the ARG'' section \llname{ignorance}
answered questions I had when reading the bits around \figref{fig:conceptual} -- upon initially reading the bit around \figref{fig:conceptual}
I was confused by the interpretation that the lack of a node implies that that individual is not represented in other parts of the ARG (as any point along any branch in an ARG represents \emph{some} individual ancestral to sample), and as a result I found the presented interpretation somewhat unnatural.
This was then eventually clarified (very eloquently) in the discussion.
Similarly, I would have appreciated the connection to parsimony earlier.
It was not completely obvious to me that recovering the unary nodes from the succinct tree sequence might not be possible in all cases, and as such that the algorithm is an attempt at inference, and furthermore, that it's optimizing some reasonable objective. Again, this point was addressed very well, but I had to wait until the discussion.
\end{point}

\reply{This is a valid point, however we think our discussion of these points was best put in the discussion section. With respect to the motivation, we were likely heavily relying on the phrase of a \emph{hypothetical ARG}, but we find this acceptable.
}

@@

\begin{point}{\revref} % p.2 l.53
 ``Notation and terminology'': it would be good to clarify at this point that parents and children in the graph correspond to ancestors and descendants in the population.
\end{point}

\reply{Good observation. We have made this change. \revref
}

\begin{point}{\revref} % p.4 l.10
 ``in practice we do repeated passes in both directions until no changes can be made'': to someone who does not have hands-on experience with the implementation of this algorithm, the reason why the algorithm needs to do several passes in both directions might not be very clear. If at each tree transition all pair of mergeable paths are merged, it would be good to explain why, after one pass in each direction, additional mergeable paths are found. Perhaps explaining when/why additional mergeable paths are found/created could also give further insight on how many passes will be needed?
\end{point}

\reply{Noted. We have written some follow up for that sentence at your suggestion. The creation of new mergable paths after a number of passes is something we thought about quite a bit, however could not come up with a good theoretical conclusion as to if there was a number of required passes before termination. We instead provide some empirical evidence to the number of required passes later in the section and \tabref{tab:edge-counts}. \revref
}

@@

\begin{point}{\revref} % p.8 l.25
 ``we simulate ARGs containing full haplotypes using msprime'': since msprime can simulate ARGs under various models, please specify here which model is used (SMC, SMC'... ?).
\end{point}

\reply{
Good suggestion; we've done this (we used ``hudson''). \revref
}

\begin{point}{}
In general, it comes natural to wonder whether the algorithms can be corrected to avoid incorrectly added spans
(e.g. \revref). % p.10 l. 5).
It would be worth to mention whether an attempt of correction has been made and to explain the difficulties/limitations involved.
\end{point}

\reply{This is a good point. It is certainly something we have thought about but we omitted additional information that would improve accuracy in place
	of simplicity of the construction. We also though our empirical results were strong without additional correction.
	We have made note of this point. \revref
}

@@

\begin{quote}
The fundamental idea of this paper is important and the worked-out
algorithm appears quite valuable. I want to see it published. However
the current version is so difficult to read that it is likely to
have a very limited audience, and appears to have some errors.
\end{quote}

\begin{quote}
I do not feel I have understood the entire paper despite several
hours' work. Some of my comments may be misunderstandings, but if
so, the text I misunderstood could probably use clarification.
\end{quote}

\begin{point}{} % (1)
The description of \algorithmref{alg:extend} and \algorithmref{alg:mergeable} is confusing.
Naming them this way implies that they are separate algorithms but
in fact ``Algorithm 2'' is the definition of a function needed by
Algorithm 1.
\end{point}

\reply{Yes, this is the case. Originally we had both \algorithmref{alg:extend} and \algorithmref{alg:mergeable} within 
	one algorithm environment. However, we discuss both the \texttt{Mergeable} and \texttt{ExtendHaplotypes} functions
	in enough detail that we thought it poignant to separate them into their own algorithms. We additionally decided to 
	make \algorithmref{alg:extend} the first algorithm as it is the main construction and the most important of the two.
}

\begin{point}{} %(2)
\algorithmref{alg:mergeable} is mentioned in an explanation a full page before it
is introduced.
\end{point}

\reply{That is true, but with our additional figures, the earliest the environment will appear is the next page following the previous figure.
	Upon consideration we think an introduction about the algorithm's conceptual construction in the section 
	and then properly defining is more easily read than the other way around.
}

\begin{point}{} %(3)
\algorithmref{alg:extend} seems to be named ExtendHaplotypes and \algorithmref{alg:mergeable} seems
to be named Mergeable: having two names for each one is not helpful.
\end{point}

\reply{We like to think of ExtendHaplotypes and Mergable as the names of the functions contained in \algorithmref{alg:extend} and \algorithmref{alg:mergeable}, respectively.
}

\begin{point}{} %(4)
``Mergeable'' is an unfortunate name for a function which does not
return a yes or no answer but instead returns the number of
new edges to be added. Also the text needs a comment about why
infinity is sometimes returned.
\end{point}

\reply{This is a credible point, however we require counting new edges that would be created by a merged path so that we can determine, out of all possible mergeable paths between trees, which to merge based on minimizing the number of newly created edges.
	For clarity, we have included a note about returning infinity in the description of \algorithmref{alg:mergeable}. \revref
}

\begin{point}{} %(5)
In \algorithmref{alg:extend} the variable $m$ means the number of new edges to be
added. In \algorithmref{alg:mergeable} that quantity is called $m_e$ and $m$ means the
number of extended haplotypes. This is confusing. Please use the
same name for a given concept in all pseudocode.
\end{point}

\reply{Good point, this change has been made. \revref
}

\begin{point}{} %(6)
Comments in the pseudocode would make it easier to follow, as could
use of more meaningful variable names (I particularly dislike $M$, $m$, $M'$
in the same algorithm).
\end{point}

\reply{Good point. We have included some comments for clarity, however since $M$, $M'$ and $m_e$ are used simply for book keeping a while loop we don't think it's necessary to use distinct variables. \revref
}

@@

\begin{point}{}
Capital $I$ is an unfortunate subscript as it is almost indistinguishable
from digit 1.
\end{point}

\reply{While we agree, within the context of our algorithm we are interested paths of inheritance 
	between the edges removed from a local tree (the out-forest) and the edges just added into the next local tree (the in-forest). 
	Because of our language use of out and in, we feel like our choice of $O$ and $I$ is reasonable.
}

@@

\begin{point}{} % (3)
I recommend that the example partial tree in \figref{fig:conceptual} be
extended to a full tree (it only needs one more node) and used in
the practical examples in the text.
\end{point}

\reply{
Good idea, but on consideration we think that adding the rest of the tree
here would make the figure too confusing
(not so much for the trees in A\&C, but more so in the haplotype diagrams B\&D).
We do now reference the figure in the text,
at \revref{} and \llname{ref_conceptual2}.
}

\begin{point}{} % (4) 
The concrete example under ``An algorithm to extend haplotypes'' is
difficult to visualize. Can it be related to \figref{fig:conceptual}? When I try to
do this, though, I have difficulty because in \figref{fig:conceptual} the right-hand
tree does not \emph{have} $v_1 \ldots v_n$, it jumps straight from $p$ to $c.$ Is
this example not mergeable, or is the description of mergeablility
missing any indication that some of the nodes it relies on may not
exist?
\end{point}

\reply{Good point. We have connected our concrete construction in that section
	 to reference \figref{fig:conceptual} to improve clarity. \revref
}

@@

\begin{point}{} % (5) 
The direction in which time is measured on the tree needs to be stated
as unfortunately the literature has both conventions. Also, the text
implies at several points that this can be used for trees without
time information, but it cannot as times are used in the algorithm.
\end{point}

\reply{We state how time is measured here \revref. We additionally do use time in our algorithm, but mainly as a way to break ties, as is the case with \eqnref{eqn:matched-spans}. This tie break will default to the first node if the ARG does not contain time information.
}

@@

\begin{point}{}
Following on from this, in \algorithmref{alg:extend}, why can no node in $u$ be at the
same time as a node in $v?$ At first I thought we were assuming that no
two nodes, other than samples, are ever at the same time (a fairly
common assumption) but then we would not need this rule. Is this just
a convenience to avoid putting two nodes on the same branch at the same
timepoint? What happens if this algorithm is attempted with inferred
ARGs that discretize time, as some inference algorithms do, so that ties
become common?
\end{point}

\reply{We do not exactly assume that all nodes in an ARG occur at unique times, however we do assume that
	any two nodes with the same descendants should be unique. 
	Additionally, in our example if $u$ and $v$ occurred at the same time and had descendant $c$, they would be nodes with the same index
	and the paths for $c$ in $T_O$ and $T_I$ would be from $c\to u=v$.
	In the case of discritized time of inferred ARGs, this is an interesting question we have not yet thought about it depth. 
	Peter do you have extra thoughts about this? 
}

@@

\begin{point}{} % (6) 
``In what follows we think of $T_2$ as the true ARG and $T_1$ as an
inferred ARG.'' This is not strongly enough stated. The measure is
asymmetrical with regards to $T_1$ and $T_2$, and will give wrong answers
if they are reversed. It is also not suitable for comparing two
ARGs where neither is the true ARG.
\end{point}

\reply{
	We think that while $\match(\T_2,\T_1)$ might not be what the user intends for our method, it can still give some valuable information between $\T_1$ and $\T_2$. We additionally discuss this when comparing inverse matched spans between $(\T_1,\T_2)$ and matched spans between $(\T_2, \T_1)$ \revref.
	While this method is developed for measuring the correctness of an inferred ARG to the true ARG, we also think of this as a function to compare two ARGs whether or not they are truly related. Also our theoretical results on this method being a metric could allow for distances in the space of all ARGs which we think is an interesting generalization within the field. 
}

@@

\begin{point}{}
In general the section on comparison of true and inferred ARGs is
difficult to follow. \figref{fig:conceptual_discrepancy} helps but appears too late in the
paper.
\end{point}

\reply{We feel like it is optimal to keep the example that computes $\operatorname{ARF}$ and $\operatorname{TPR}$
	until after the are defined, and hence why it is at the end of the section. We have however, heard your point and
	so we have expanded our example to give more clarity. \revref
}

@@

\begin{point}{} % (7) 
In \figref{fig:conceptual_discrepancy} it is unfortunate to use to use an example where
the forward and inverse matches are the same. I would also appreciate
a brief explanation of why it's 46 and 47 in the denominator--I figured
it out but it's less than obvious. Why do sample nodes count, as they
cannot vary among trees? Doesn't this just add bookkeeping effort?
It's otherwise a very helpful figure though.
\end{point}

\reply{Good point. It is unfortunate how the math works out, but we did construct the example as such to discern that the matching for node 4 requires a max to be taken over shared spans of its two possible matches 4 and 9. To show some variety, we additionally will compute $\match(T2,T1)$ to show some additional variety and hopefully give more clarity to the methods themselves \revref. Additionally we note that we are matching spans for all nodes in a pair of ARGs including the samples however they trivially match to themselves by construction (this is not a hassle for bookkeeping). We also must have that the two ARGs being compared has the same number of sample nodes.
}

@@

\begin{point}{} % (8) 
I cannot clearly distinguish the three purplish colors in \figref{fig:speed_and_edges}.
\end{point}

\reply{Thank you for this feedback. We have changed \figref{fig:speed_and_edges} and the subsequent supplementary figures (\figref{fig:speed_and_edges_constant}, \figref{fig:speed_and_edges_other}, \figref{fig:speed_and_edges_other_constant}) to have more color variety.
}

@@

\begin{point}{} % (9) 
The interpretation of \figref{fig:node-spans} is completely unclear to me. I
think it needs more explanatory text. I have no idea what point is being
made by these diagrams.
\end{point}

\reply{Thanks for the feedback. We have added some additional feedback \revref{} and the caption of \figref{fig:node-spans}.
}

### 2025-05-16 paper/writeup.tex
reviews (mostly) complete

Consider the (small portion) of a hypothetical ARG in \figref[A]{fig:conceptual}.
On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
coalesce into a small subtree: 1 and 2 find a common ancestor in ancestral node 3,
which finds a common ancestor with node 0 in ancestral node 4.
On the next portion of the genome (right-hand tree), sample node 2 has a different ancestor.
This seems reasonable, and a method that infers trees separately on each portion of the genome
could not be expected to produce anything different.
However, the example becomes more complicated once we consider what these local genealogies imply about haplotype inheritance.
\figref[B]{fig:conceptual} shows the implied inheritance of haplotypes,
with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
Here, sample node 2 has inherited the chunk of haplotype labeled $L$ from ancestral node 4 via 3,
and the haplotype to the right of this from some other node (and so doesn't carry haplotype $R$).
On the other hand, sample node 1 has inherited \emph{both} haplotypes $L$ and $R$
from ancestral node 4, but the trees imply that only haplotype $L$ is inherited via ancestral node 3.
This implies -- if taken literally -- that there must have been a recombination event
at some point between node 1 and node 4 that separated the $L$ and $R$ haplotypes,
and then these two ancestral (and nonoverlapping) haplotypes coalesced together in ancestral node 4.
Although this is possible, it seems unlikely --
a more parsimonious explanation is depicted in \figref[C]{fig:conceptual},
in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
(and there is a recombination somewhere between node 3 and node 2).
This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
which is depicted in \figref[D]{fig:conceptual} --
and so node 3 has become unary in this tree.
Note that the more parsimonious ARG also includes fewer edges:
the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref[B]{fig:conceptual}
have been reduced to the two edges
$4 \to 3$ and $3 \to 1$ in \figref[D]{fig:conceptual}.

@@

We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
to take advantage of the % extensive, stable, and well-documented set of open-source 
tools available in \tskit{} \citep{tskit}, 
and our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived.
Some nodes are \emph{samples}, meaning that they represent genome sequences available as data.
% Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$, \revpoint{2}{1}
over a segment of genome $[\ell_e, r_e)$.
% (The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Nodes represent (portions of) ancestral haplotypes, and so
we will use the terms interchangeably.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:extending_diagram},
we do this by identifying paths of inheritance that are shared across a sequence of local trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
For instance, in \figref{fig:conceptual} take $p=4$ and $c=1$,
so that $u_1=3$ (and $m=1)$ and there are no $v$ (so $n=0$); \revpoint{3}{11}
and so as shown in \figref[B\&D]{fig:conceptual},
we might extend node 3 over this entire segment.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$, for some $i$ and $j$.
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

@@

\algorithmref{alg:extend} simplifies the full algorithm implemented in software % in \tskit{}
in several ways for the sake of clarity --
for instance, the bookkeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
in practice we do repeated passes in both directions until no changes can be made. \revpoint{2}{2}
The main step that is omitted is a description of the \texttt{merge} operation,
which performs the actual extending of haplotypes.
This algorithm is essentially the same as \texttt{mergeable} in \algorithmref{alg:mergeable},
except with additional bookkeeping.
Roughly speaking, the algorithm
traverses up from the shared base node $c$,
doing the appropriate operations to insert the nodes along the path found in $T_k$
into the path in $T_{k+1}$.
To do this, some edges that end at $a_k$ will be extended to end at $a_{k+1}$;
some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
and some entirely new edges may be added, as in \figref{fig:extending_diagram}.
Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
need to be updated.

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Extend haplotypes. Given an ARG $\T$ with $N$ trees $T_1, \ldots, T_N$
        for which edges $O_k$ are removed to transition from $T_k$ to $T_{k+1}$,
        identify and merge all mergeable paths (see text).
        Each child node $c_e$ of each removed edge $e$ is checked to see if it
        is at the base of two mergeable paths;
        paths that add fewer new edges are merged first
        because \texttt{Mergeable} returns the number of new edges required
        to merge the two paths, and $M$ is always less than or equal to
        the minimum number of new edges across all nodes in the current tree transition.
    }
	\label{alg:extend}
	\SetKwFunction{mergeable}{Mergeable}
	\SetKwFunction{merge}{Merge}
	\SetKwFunction{extend}{ExtendHaplotypes}
	\DontPrintSemicolon
    \Fn{\extend($\T$)}{
        \For{$k$ in $1 \ldots N$}{
            \text{Set } $M = 0$ \text{ and }
            $M' = \infty$.\;
            \While{$M < \infty$}{
                \For{$e \in O_k$}{
                    \text{Set } $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\;
                    \eIf{$m \le M$}{
                        \merge($c_e$, $T_k$, $T_{k+1}$)\;
                    }{
                        \text{Set } $M' = \min(m, M)$. \;
                    }
                }
                \text{Set } $M = M'$ \text{ and }
                $M' = \infty$ .\;
            }
        }
    }
\end{algorithm2e}

>>>>

Consider the (small portion of a) hypothetical ARG in \figref[A]{fig:conceptual}.
On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
coalesce into a small subtree: 1 and 2 find a common ancestor in ancestral node 3,
which finds a common ancestor with node 0 in ancestral node 4.
On the next portion of the genome (right-hand tree), sample node 2 has a different ancestor.
This seems reasonable, and a method that infers trees separately on each portion of the genome
could not be expected to produce anything different.
However, the example becomes more complicated once we consider what these local genealogies imply about haplotype inheritance.
\figref[B]{fig:conceptual} shows the implied inheritance of haplotypes,
with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
Here, sample node 2 has inherited the chunk of haplotype labeled $L$ from ancestral node 4 via 3,
and the haplotype to the right of this from some other node (and so doesn't carry haplotype $R$).
On the other hand, sample node 1 has inherited \emph{both} haplotypes $L$ and $R$
from ancestral node 4, but the trees imply that only haplotype $L$ is inherited via ancestral node 3.
This implies -- if taken literally -- that there must have been a recombination event
at some point between node 1 and node 4 that separated the $L$ and $R$ haplotypes,
and then these two ancestral (and nonoverlapping) haplotypes coalesced together in ancestral node 4.
Although this is possible, it seems unlikely --
a more parsimonious explanation is depicted in \figref[C]{fig:conceptual},
in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
(and there is a recombination somewhere between node 3 and node 2).
This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
which is depicted in \figref[D]{fig:conceptual} --
and so node 3 has become unary in this tree.
Note that the more parsimonious ARG also includes fewer edges:
the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref[B]{fig:conceptual}
have been reduced to the two edges
$4 \to 3$ and $3 \to 1$ in \figref[D]{fig:conceptual}.

@@

We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
to take advantage of the % extensive, stable, and well-documented set of open-source 
tools available in \tskit{} \citep{tskit}, 
and our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived. \revpoint{3}{13}
Some nodes are \emph{samples}, meaning that they represent genome sequences available as data.
% Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between ancestors and their descendants within the population.
The ancestor and descendant are represented by the parent node $p_e$ and child node $c_e$ respectively \revpoint{2}{1}
over a segment of genome $[\ell_e, r_e)$.
% (The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Nodes represent (portions of) ancestral haplotypes, and so
we will use the terms interchangeably.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:extending_diagram},
we do this by identifying paths of inheritance that are shared across a sequence of local trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
For instance, in \figref{fig:conceptual} take $p=4$ and $c=1$,
so that $u_1=3$ (and $m=1)$ and there are no $v$ (so $n=0$); \revpoint{3}{11}\revpoint{3}{12}
and so as shown in \figref[B\&D]{fig:conceptual},
we might extend node 3 over this entire segment.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$, for some $i$ and $j$.
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

@@

\algorithmref{alg:extend} simplifies the full algorithm implemented in software % in \tskit{}
in several ways for the sake of clarity --
for instance, the bookkeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
in practice we do repeated passes in both directions until no changes can be made. \revpoint{2}{2}
We require repeated passes due to the fringe cases of some paths not being mergeable until 
additional structure is added by a future pass of the algorithm. 
We suspect that most of these cases occur for paths with large time differences between the parent and child
nodes, or when a lot of nodes in a path are coalescent, however it is not something we thoroughly investigated
in the theoretical setting.
We instead give some empirical results on the number of passes for the algorithm below.

The main step that is omitted is a description of the \texttt{Merge} operation,
which performs the actual extending of haplotypes.
This algorithm is essentially the same as \texttt{Mergeable} in \algorithmref{alg:mergeable},
except with additional bookkeeping.
Roughly speaking, the algorithm
traverses up from the shared base node $c$,
doing the appropriate operations to insert the nodes along the path found in $T_k$
into the path in $T_{k+1}$.
To do this, some edges that end at $a_k$ will be extended to end at $a_{k+1}$;
some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
and some entirely new edges may be added, as in \figref{fig:extending_diagram}.
Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
need to be updated.

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Extend haplotypes. Given an ARG $\T$ with $N$ trees $T_1, \ldots, T_N$
        for which edges $O_k$ are removed to transition from $T_k$ to $T_{k+1}$,
        identify and merge all mergeable paths (see text).
        Each child node $c_e$ of each removed edge $e$ is checked to see if it
        is at the base of two mergeable paths;
        paths that add fewer new edges are merged first
        because \texttt{Mergeable} returns the number of new edges required
        to merge the two paths, and $M$ is always less than or equal to
        the minimum number of new edges across all nodes in the current tree transition.
    }
	\label{alg:extend}
	\SetKwFunction{mergeable}{Mergeable}
	\SetKwFunction{merge}{Merge}
	\SetKwFunction{extend}{ExtendHaplotypes}
	\DontPrintSemicolon
    \Fn{\extend($\T$)}{
    	\tcc{For pairs of trees in the ARG, find mergeable paths.}
        \For{$k$ in $1 \ldots N$}{
            \text{Set } $M = 0$ \text{ and }
            $M' = \infty$.\;
            \While{$M < \infty$}{
                \For{$e \in O_k$}{
                    \text{Set } $m_e$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\tcc*[r]{Count the number of new edges to be merged.}
                    \eIf{$m_e \le M$}{
                        \merge($c_e$, $T_k$, $T_{k+1}$)\tcc*[r]{Merge extended haplotypes from $T_k$ into $T_{k+1}$.}
                    }{
                        \text{Set } $M' = \min(m_e, M)$.\tcc*[r]{Bookkeep $M$ to determine an ordering for path merging.}
                    }
                }
                \text{Set } $M = M'$ \text{ and }
                $M' = \infty$ .\;
            }
        }
    }
\end{algorithm2e}
\revpoint{3}{5}
\revpoint{3}{4}
\revpoint{3}{6}
\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Given a node $c$, trees $T_O$ and $T_I$,
        and sub-forests $F_O$ and $F_I$
        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
        check to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
        If the paths are mergeable then this returns the number of new edges
        that would be added by extending the path from $T_O$ to $T_I$;
        otherwise, this returns $\infty$.
        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
        in the set of edges to be removed and added, respectively
        (i.e., in $F_O$ and $F_I$).
        The variable $m_e$ will record the number of new edges to be added,
        and $m$ will record the number of extended haplotypes.
        Note that if no haplotypes can be extended ($m=0$), we return $m_e$ as $\infty$
        so that the path is not merged in \algorithmref{alg:extend}.
    }
	\label{alg:mergeable}
	\SetKwFunction{mergeable}{Mergeable}
	\DontPrintSemicolon
    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
    	\tcc{We construct paths beginning at $c$ and traverse through inheritance in $T_O$ and $T_I$.}
        \text{Let } $p_i = P_I[c]$, 
        $t_i = t[p_i]$,
        $p_o = P_O[c]$,
        $t_o = t[p_o]$,
        \text{ and }$m_e = m = 0$.\;
        \While{\tn{True}}{
        	\tcc{States to determine if paths can still be constructed backward in time.}
            \text{Set } $y_i = (p_i \neq NULL)
                \ \&\ (p_i \notin T_O)
                \ \&\ (t_i < t_o)$ \;
            \text{and } $y_o = (p_o \neq NULL)
                \ \&\ (p_o \notin T_I)
                \ \&\ (t_o < t_i)$ \;
